### 2018-02-23 pubs/itp-2018/sections/metis-language-and-proofs.tex
ITP 2018: Initial commit.
[ paper ] removed many examples and extra content.

\begin{myexamplenum}
  For instance, let us express the problem
  $p\, ⊢ ¬ (p ∧ ¬ p) ∨ (q ∧ ¬ q)$ in \TPTP syntax. We begin by
  declaring the $p$ axiom using the \verb!axiom!  keyword. Next, we
  include the expected conclusion using the \verb!conjecture! keyword.

\begin{verbatim}
  fof(h, axiom, p).
  fof(goal, conjecture, ~ ((p & ~ p) | (q & ~ q))).
\end{verbatim}

\end{myexamplenum}

@@

\begin{myexamplenum}

In the script below, \strip is the name of the inference.
It has no arguments and derives from one parent node named  \verb!goal!. The
result of this inference when it applies to the \verb!goal! formula is
\verb!p!.

\begin{verbatim}
  fof(subgoal_0, plain, p, inference(strip, [], [goal])).
\end{verbatim}
\end{myexamplenum}

@@

\begin{verbatim}
  fof(premise, axiom, p).
  fof(goal, conjecture, p).
  fof(s₀, p, inf(strip, goal)).
  fof(neg₀, ¬ p, inf(negate, s₀)).
  fof(n₀, ¬ p, inf(canonicalize, neg₀)).
  fof(n₁, p, inf(canonicalize, premise)).
  fof(n₂, ⊥, inf(simplify, [n₀, n₁]))
  cnf(r₀, ⊥, inf(canonicalize, n₂)).
\end{verbatim}

@@

A derivation generated by \Metis encodes a natural
deduction proof, Fig.~\ref{fig:metis-example} is an example of such kind of proof. With the inference rules in
Fig.~\ref{tab:agda-metis-table}
as the only valid deduction steps, \Metis
attempts to prove conjectures by refutation (\ie,
\emph{falsium} in the root of the \TSTP derivation).

@@

\begin{figure}[!ht]
\centering
  \begin{bprooftree}\tt
    \AxiomC{}
    \RightLabel{negate}
    \UnaryInfC{$\neg p$}
    \RightLabel{strip}
    \UnaryInfC{$\neg p$}
    \AxiomC{}
    \RightLabel{axiom}
    \UnaryInfC{$p$}
    \RightLabel{canonicalize}
    \UnaryInfC{$p$}
    \RightLabel{simplify}
    \BinaryInfC{$⊥$}
    \RightLabel{canonicalize}
    \UnaryInfC{$⊥$}
  \end{bprooftree}
  \caption{\Metis derivation tree of Example~\ref{fig:metis-example-tree}.}
  \label{fig:metis-example}
\end{figure}

@@

We first present the \strip inference rule since it is the rule that appears
first after each conjecture. The other rules are sorted mainly follow their
level of complexity of their definitions and the formalization presented in
Section~\ref{ssec:emulating-inferences}. Some inference rules depend on the
formalizations of other rules. For instance, the \simplify rule and the
\clausify rule need theorems developed for the \canonicalize rule. The
\canonicalize rule needs theorems developed in the \resolve section. But the
\resolve rule depends on the \conjunct rule.

@@

\caption{\Metis inference rules.}
  \begin{center}
  {\renewcommand{\arraystretch}{1.6}%
    \label{tab:agda-metis-table}
    \begin{tabular}{|@{\hspace{2mm}}l@{\hspace{4mm}}l@{\hspace{2mm}}c@{\hspace{2mm}}|}
    \hline
    \textbf{\Metis rule} & \textbf{Purpose} &\textbf{Theorem number}\\ \hline

>>>>

@@

@@

\begin{verbatim}
  fof(premise, axiom, p).
  fof(goal, conjecture, p).
  fof(s₀, inf(strip, goal)).
  fof(neg₀, ¬ p, inf(negate, s₀)).
  fof(n₀, ¬ p, inf(canonicalize, neg₀)).
  fof(n₁, p, inf(canonicalize, premise)).
  fof(n₂, ⊥, inf(simplify, [n₀, n₁]))
  cnf(r₀, ⊥, inf(canonicalize, n₂)).
\end{verbatim}

@@

A derivation generated by \Metis encodes a natural
deduction proof. With the inference rules in
Fig.~\ref{tab:agda-metis-table}
as the only valid deduction steps, \Metis
attempts to prove conjectures by refutation (\ie,
\emph{falsium} in the root of the \TSTP derivation).

@@

@@

We first present the \strip inference rule since it is the rule that appears
first after each conjecture. The other rules are sorted mainly follow their
level of complexity of their definitions and the formalization presented in
Section~\ref{ssec:emulating-inferences}. Some inference rules depend on the
formalizations of other rules. For instance, the \simplify rule and the
\clausify rule need theorems developed for the \canonicalize rule. The
\canonicalize rule needs theorems developed in the \resolve formalization,
and \resolve depends on the \conjunct rule.

@@

\caption{\Metis inference rules.}
  \begin{center}
  {\renewcommand{\arraystretch}{1.6}%
    \label{tab:agda-metis-table}
    \begin{tabular}{|@{\hspace{2mm}}l@{\hspace{4mm}}l@{\hspace{2mm}}c@{\hspace{2mm}}|}
    \hline
    \textbf{Rule} & \textbf{Purpose} &\textbf{Theorem number}\\ \hline

### 2018-02-23 pubs/itp-2018/sections/proof-reconstruction/strip.tex
[ paper ] removed many examples and extra content.

\begin{proof}
The proof is by induction on the cases defined by the outcome of the
$\fuh_{1}$ function.
\begin{itemize}
  \item If $n = 0$, by definition in~\eqref{eq:uh-structured}
        we conclude $Γ ⊢ φ$.
  \item For $n \geq 1$, we use induction on the structure of the first
        argument.
\vskip 2mm
\begin{itemize}
\item Case $φ ≡ φ₁ ⇒ (φ₂ ⇒ φ₃)$.
\begin{equation*}
  \begin{bprooftree}
  \AxiomC{$Γ ⊢ \fuh_1~(φ₁ ⇒ (φ₂ ⇒ φ₃))~(\suc~n)$}
  \RightLabel{by~(\ref{eq:uh-structured})}
  \UnaryInfC{$Γ ⊢ \fuh_1~((φ₁ ∧ φ₂) ⇒ φ₃)~n$}
  \RightLabel{by~ind.~hyp.}
  \UnaryInfC{Γ ⊢ (φ₁ ∧ φ₂) ⇒ φ₃}
  \RightLabel{∧⇒\textsf{-to-}⇒⇒}
  \UnaryInfC{Γ ⊢ φ₁ ⇒ (φ₂ ⇒ φ₃)}
  \end{bprooftree}
\end{equation*}
using the following theorem proved in~\cite{AgdaProp},
  \begin{equation*}
  \texttt{∧⇒\textsf{-to-}⇒⇒}\ :\  Γ ⊢ (φ₁ ∧ φ₂) ⇒ φ₃ → Γ ⊢ φ₁ ⇒ (φ₂ ⇒ φ₃).
  \end{equation*}

\item Case $φ ≡ φ₁ ⇒ (φ₂ ∧ φ₃)$.
\begin{equation*}
  (\mathcal{D}_1)
  \begin{bprooftree}
    \AxiomC{$Γ ⊢ \fuh_1~(φ₁ ⇒ (φ₂ ∧ φ₃))~(\suc~n)$}
    \RightLabel{by~(\ref{eq:uh-structured})}
    \UnaryInfC{$Γ ⊢ \fuh_1~(φ₁ ⇒ φ₂)~n ∧ \fuh_1~(φ₁ ⇒ φ₃)~n$}
    \RightLabel{∧-proj₁}
    \UnaryInfC{$Γ ⊢ \fuh_1~(φ₁ ⇒ φ₂)~n$}
    \RightLabel{by~ind.~hyp.}
    \UnaryInfC{$Γ ⊢ φ₁ ⇒ φ₂$}
  \end{bprooftree}
\end{equation*}

@@

∙~Case $φ ≡ ¬ (¬ φ₁) $.
\begin{equation*}
  \begin{bprooftree}
 \AxiomC{$Γ ⊢ \fstrip₁~(¬(¬φ₁))~(\suc~n)$}
  \RightLabel{by~(\ref{eq:strip-fixed})}
  \UnaryInfC{$Γ ⊢ \fuh~(\fstrip₁~φ₁~n)$}
  \RightLabel{Lemma~\ref{lem:inv-uh-lem}}
  \UnaryInfC{$Γ ⊢ \fstrip₁~φ₁~n$}
  \RightLabel{by~ind.~hyp.}
  \UnaryInfC{$Γ ⊢ φ₁$}
  \end{bprooftree}
\end{equation*}

∙~Case $φ ≡ φ₁ ∧ φ₂$. We prove $Γ ⊢ φ₁$ and $Γ ⊢ φ₂$.
From the conjunction of φ₁ and φ₂, the expected result follows.

\begin{equation*}
(\mathcal{D})\hspace{3mm}
  \begin{bprooftree}
  \AxiomC{$Γ ⊢ \fstrip₁~(φ₁ ∧ φ₂)~(\suc\,\textit{n})$}
  \RightLabel{by~(\ref{eq:strip-fixed})}
  \UnaryInfC{$Γ ⊢ \fuh~(\fstrip₁~φ₁~n) ∧ \fuh~(φ₁ ⇒ \fstrip₁~φ₂~n)$}
  \RightLabel{∧-proj₁}
  \UnaryInfC{$Γ ⊢ \fuh~(\fstrip₁~φ₁~n)$}
  \RightLabel{Lemma~\ref{lem:inv-uh-lem}}
  \UnaryInfC{$Γ ⊢ \fstrip₁~φ₁~n$}
  \RightLabel{by~ind.~hyp.}
  \UnaryInfC{$Γ ⊢ φ₁$}
    \end{bprooftree}
\end{equation*}

\begin{equation*}
  \begin{bprooftree}
  \AxiomC{$\mathcal{D}$}
  \UnaryInfC{Γ ⊢ φ₁}
  \AxiomC{$Γ ⊢ \fstrip₁~(φ₁ ∧ φ₂)~(\suc~n)$}
  \RightLabel{by~(\ref{eq:strip-fixed})}
  \UnaryInfC{$Γ ⊢ \fuh~(\fstrip₁~φ₁~n) ∧ \fuh~(φ₁ ⇒ \fstrip₁~φ₂~n)$}
  \RightLabel{$∧$-proj$₂$}
  \UnaryInfC{$Γ ⊢ \fuh~(φ₁ ⇒ \fstrip₁~φ₂~n)$}
  \RightLabel{Lemma~\ref{lem:inv-uh-lem}}
  \UnaryInfC{$Γ ⊢ φ₁ ⇒ \fstrip₁~φ₂~n$}
  \RightLabel{$⇒$-elim}
  \BinaryInfC{$Γ ⊢ \fstrip₁~φ₂~n$}
  \RightLabel{by~ind.~hyp.}
  \UnaryInfC{$Γ ⊢ φ₂$}
  \end{bprooftree}
\end{equation*}

@@

Since \Metis proves a conjecture by refutation,
to prove each subgoal, \Metis assumes the negation of it
by using the \negate rule after the \strip inference application
that introduce such a subgoal.

\begin{myexamplenum}
In the following \TSTP derivation, note that the three subgoals
\verb!s₁!, \verb!s₂! and \verb!s₃! are assumed by the \negate
rule in \verb!neg₁!, \verb!neg₂! and  \verb!neg₃! respectively.

\begin{verbatim}
  fof(goal, conjecture, p ∧ r ∧ q).
  fof(s₁, p, inf(strip, goal)).
  fof(s₂, p ⇒ r, inf(strip, goal)).
  fof(s₃, (p ∧ r) ⇒ q, inf(strip, goal)).
  fof(neg₁, ¬ p, inf(negate, s₁)).
  ...
  fof(neg₂, ¬ (p ⇒ r), inf(negate, s₂)).
  ...
  fof(neg₃, ¬ ((p ∧ r) ⇒ q), inf(negate, s₃)).
\end{verbatim}
\end{myexamplenum}

>>>>

\begin{proof}
The proof is by induction on the cases defined by the outcome of the
$\fuh_{1}$ function.
\begin{itemize}
  \item If $n = 0$, by definition in~\eqref{eq:uh-structured}
        we conclude $Γ ⊢ φ$.
  \item For $n \geq 1$, we use induction on the structure of the first
        argument.
\vskip 2mm
\begin{itemize}
\item Case $φ ≡ φ₁ ⇒ (φ₂ ∧ φ₃)$.
\begin{equation*}
  (\mathcal{D}_1)
  \begin{bprooftree}
    \AxiomC{$Γ ⊢ \fuh_1~(φ₁ ⇒ (φ₂ ∧ φ₃))~(\suc~n)$}
    \RightLabel{by~(\ref{eq:uh-structured})}
    \UnaryInfC{$Γ ⊢ \fuh_1~(φ₁ ⇒ φ₂)~n ∧ \fuh_1~(φ₁ ⇒ φ₃)~n$}
    \RightLabel{∧-proj₁}
    \UnaryInfC{$Γ ⊢ \fuh_1~(φ₁ ⇒ φ₂)~n$}
    \RightLabel{by~ind.~hyp.}
    \UnaryInfC{$Γ ⊢ φ₁ ⇒ φ₂$}
  \end{bprooftree}
\end{equation*}

@@

@@

Since \Metis proves a conjecture by refutation,
to prove each subgoal, \Metis assumes the negation of it
by using the \negate rule after the \strip inference application
that introduces such a subgoal.
